---
title: "ggmice"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggmice}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE, 
  message = FALSE,
  fig.width = 6, 
  fig.height = 6
)
```

```{r setup, include = FALSE}
# set-up environment
# library(shinymice)
library(tidyverse)
`%nin%` <- Negate(`%in%`)

# example dataset
dat <- mice::boys
# example mids object
imp <- mice::mice(dat, printFlag = FALSE)

# define mice theme
theme_mice <- function() {
  # move this outside of the function to unnecessary avoid re-running
  ggplot2::update_geom_defaults("point", list(
    shape = 21,
    size = 1.9,
    stroke = 1.1,
    alpha = 0.5
  ))
  ggplot2::update_geom_defaults("boxplot", list(
    size = 1
  ))
  ggplot2::update_geom_defaults("line", list(
    size = 1
  ))
  # the actual theme settings
  theme <- list(
    ggplot2::theme_classic(),
    ggplot2::theme(legend.position = "top"),
    ggplot2::scale_color_manual(
      values = c(
        "observed" = mice:::mdc(1),
        "missing" = mice:::mdc(2),
        "imputed" = mice:::mdc(2)
      )
    ),
    ggplot2::scale_fill_manual(
      values = c(
        "observed" = mice:::mdc(1),
        "missing" = mice:::mdc(2),
        "imputed" = mice:::mdc(2)
      )
    ),
    ggplot2::scale_size_manual(
      values = c("observed" = 1,
      "imputed" = 0.5)
),
    ggplot2::labs(color = NULL, fill = NULL, size = NULL)
  )
  return(theme)
}
```

## Plot incomplete data

We can evaluate the missingness in the incomplete data using two plotting devices:

- histogram

- scatterplot


```{r}
# histogram/density plot faceted instead of dodged
plot_conditional <- function(dat, x, z) {
  # define graphing elements to add to plot
  if (is.numeric(dat[[x]])) { # for continuous variables
    geom <-
      ggplot2::geom_histogram()
  } else { # for categorical variables
    geom <-
      ggplot2::geom_bar()
  }
  
  # preprocess the data
  dat[!is.na(dat[[x]]), ] %>%
    dplyr::mutate(conditional = factor(
      is.na(.data[[z]]),
      levels = c(FALSE, TRUE),
      labels = c("observed", "missing")
    )) %>%
    # plot
    ggplot2::ggplot(ggplot2::aes(
        x = .data[[x]],
        color = conditional,
        fill = "white"
      )) +
    geom +
    # split by conditional variable
    ggplot2::facet_wrap(
      ~ conditional,
      ncol = 1,
      scales = "free_y",
      labeller = ggplot2::labeller(conditional = c(
        paste(z, "observed"), paste(z, "missing")
      ) %>% setNames(c(
        "observed", "missing"
      )))
    ) +
    # style
    theme_mice() +
    theme(legend.position = "none",
          strip.background = element_rect(size = 0.5))
}

# test
plot_conditional(dat, x = "age", z = "hc")
plot_conditional(dat, x = "reg", z = "hc")
```


```{r}
# scatterplot with NAs
plot_bivariate <- function(dat, x, y) {
  # define graphing elements to add to plot later
  # set NA value and scale for variable x
  if (is.numeric(dat[[x]])) {
    NA_x <-
      min(dat[[x]], na.rm = TRUE) - .1 * diff(range(dat[[x]], na.rm = TRUE))
    shade_x <-
      min(dat[[x]], na.rm = TRUE) - .05 * diff(range(dat[[x]], na.rm = TRUE))
    scale_x <- NULL
  } else {
    NA_x <- "NA"
    shade_x <- 1.5
    scale_x <-
      ggplot2::scale_x_discrete(limits = c("NA", levels(dat[[x]])))
  }
  
  # set NA value and scale for variable y
  if (is.numeric(dat[[y]])) {
    NA_y <-
      min(dat[[y]], na.rm = TRUE) - .1 * diff(range(dat[[y]], na.rm = TRUE))
    shade_y <-
      min(dat[[y]], na.rm = TRUE) - .05 * diff(range(dat[[y]], na.rm = TRUE))
    scale_y <- NULL
  } else {
    NA_y <- "NA"
    shade_y <- 1.5
    scale_y <-
      ggplot2::scale_y_discrete(limits = c("NA", levels(dat[[y]])))
  }
  
  # add geom for NAs in the x space
  if (any(is.na(dat[[x]]))) {
    geom_x <- ggplot2::geom_point(
      position = ggplot2::position_jitter(width = 0.2, height = 0.2),
      color = mice:::mdc(2),
      shape = 4,
      mapping = ggplot2::aes(x = NA_x, y = .data[[y]]),
      data = dat[is.na(dat[[x]]),]
    )
  } else {
    geom_x <- NULL
  }
  
  # add geom for NAs in the y space
  if (any(is.na(dat[[y]]))) {
    geom_y <- ggplot2::geom_point(
      position = ggplot2::position_jitter(width = 0.2, height = 0.2),
      color = mice:::mdc(2),
      shape = 4,
      mapping = ggplot2::aes(x = .data[[x]], y = NA_y),
      data = dat[is.na(dat[[y]]),]
    )
  } else{
    geom_y <- NULL
  }
  
  # add geom for NAs in the x-y space
  if (any(is.na(dat[[x]]) & is.na(dat[[y]]))) {
    geom_xy <- ggplot2::geom_point(
      position = ggplot2::position_jitter(width = 0.2, height = 0.2),
      color = mice:::mdc(2),
      shape = 4,
      mapping = ggplot2::aes(x = NA_x, y = NA_y),
      data = dat[is.na(dat[[x]]) & is.na(dat[[y]]),]
    )
  } else{
    geom_xy <- NULL
  }
  
  # plot
  p <- dat %>% ggplot2::ggplot() +
    # plot observed datapoints in x-y space
    ggplot2::geom_point(
      ggplot2::aes(x = .data[[x]], y = .data[[y]]),
      position = ggplot2::position_jitter(width = 0.2, height = 0.2),
      color = mice::mdc(1)
    ) +
    # add area for NAs in the x space
    ggplot2::annotate(
      "rect",
      xmin = -Inf,
      xmax = shade_x,
      ymin = -Inf,
      ymax = Inf,
      alpha = 0.1
    ) +
    # add area for NAs in the y space
    ggplot2::annotate(
      "rect",
      xmin = -Inf,
      xmax = Inf,
      ymin = -Inf,
      ymax = shade_y,
      alpha = 0.1
    ) +
    # add NAs 
    geom_x +
    geom_y +
    geom_xy +
    # style graph
    theme_mice() +
    scale_x +
    scale_y 
  
  # output
  return(p)
}

# test
plot_bivariate(dat, x = "reg", y = "hc")
plot_bivariate(dat, x = "hgt", y = "hc") 
```

## Plot missingness

- md pattern

- fluxplot

First look at the missing data pattern

```{r fig.height=7.5, fig.width=6}
# make ggplot version of md pattern plot
plot_md_pattern <- function(dat) {
  # get md pattern and store additional info
  pat <- mice::md.pattern(dat)
  vrb <- colnames(pat)[-ncol(pat)]
  colnames(pat) <- c(vrb, "n_vrb_inc")
  n_pat_obs <- as.numeric(rownames(pat))[-nrow(pat)]
  n_vrb_inc <- as.numeric(pat[, ncol(pat)])[-nrow(pat)]
  n_val_mis <- as.numeric(pat[nrow(pat), ])[-ncol(pat)]
  
  # make the pattern tidy
  long_pat <- pat[-nrow(pat), ] %>%
    cbind(., n_pat_obs, pat_nr = 1:nrow(.)) %>%
    as.data.frame() %>%
    tidyr::pivot_longer(cols = all_of(vrb),
                        names_to = "vrb",
                        values_to = "obs") %>%
    cbind(., n_val_mis)
  
  # plot the md pattern
  p <- long_pat %>%
    ggplot2::ggplot() +
    ggplot2::geom_tile(ggplot2::aes(
      x = vrb,
      y = pat_nr,
      fill = as.factor(obs),
      group = n_vrb_inc
    ),
    color = "black") +
    # set axes
    ggplot2::scale_x_discrete(limits = vrb,
                              position = "bottom",
                              labels = as.character(n_val_mis)) +    
    ggplot2::scale_y_reverse(
      breaks = 1:max(long_pat$pat_nr),
      labels = as.character(n_pat_obs),
      expand = c(0.01, 0.01),
      sec.axis = ggplot2::dup_axis(
        labels = as.character(n_vrb_inc),
        name = "Incomplete variables (per pattern)")
    ) +
    # add labels
    ggplot2::labs(x = "Incomplete cases (per variable)",
                  y = "Pattern frequency") +
    ggplot2::geom_text(aes(x = vrb, y = -Inf, label = vrb),
              data = long_pat,
              vjust = -0.5) +
    ggplot2::coord_cartesian(clip = "off") +
    # add styling
    ggplot2::theme(
      legend.position = "none",
      plot.margin = ggplot2::margin(25, 10, 10, 10, unit = "pt"),
      axis.title.y.right = ggplot2::element_text(margin = ggplot2::margin(l = 10)),
      panel.grid = ggplot2::element_blank()
    ) +
    ggplot2::scale_fill_manual(values = c("1" = mice:::mdc(1), "0" = mice:::mdc(2))) 
  
  return(p)
}

# test
plot_md_pattern(dat)
```

Then look at the in- and outflux

```{r}
# plot in- and outflux of incomplete data
plot_flux <- function(dat){
  flx <- mice::flux(dat)
  p <- flx %>% ggplot2::ggplot() +
    ggplot2::geom_text(
      ggplot2::aes(x = influx, y = outflux, label = rownames(flx))
    ) +
    geom_abline(intercept = 1, slope = -1, linetype = "dashed") +
    lims(x = c(0,1), y = c(0,1)) +
    theme_classic()
  return(p)
}
# test
plot_flux(dat)
```


## Plot convergence

```{r}

# make chain means and variances tidy
preprocess_thetas <- function(imp) {
  d <- imp$chainMean %>%
    dplyr::na_if(., "NaN") %>%
    as.data.frame() %>%
    dplyr::mutate(var = row.names(.), theta = "Chain means") %>%
    rbind(
      .,
      imp$chainVar %>%
        sqrt() %>%
        as.data.frame(.) %>%
        dplyr::mutate(var = row.names(.), theta = "Chain standard deviations")
    ) %>%
    tidyr::pivot_longer(-c(var, theta)) %>%
    cbind(.it = as.integer(1:imp$iteration),
          .imp = as.factor(rep(1:imp$m, each = imp$iteration)))
  return(d)
}
# test
trace_dat <- preprocess_thetas(imp)

trace_one_variable <- function(d, x) {
  # plot one variable
  p <- d[d$var == x,] %>%
    ggplot2::ggplot() +
    ggplot2::geom_line(ggplot2::aes(x = .it, y = value, color = .imp)) +
    ggplot2::facet_wrap( ~ theta, scales = "free", ncol = 1) +
    ggplot2::theme_classic() +
    ggplot2::theme(
      strip.background = element_rect(size = 0.5)#,
      #strip.text = element_text(face = "bold", size = 11),
      #legend.position = "bottom"
    ) +
    ggplot2::labs(x = "Iteration",
                  y = paste0(x),
                  color = "Imputation")
  # show message to user when variable is completely observed
  if (all(is.na(p$data$value))) {
    p <-
      p + ggplot2::geom_text(ggplot2::aes(
        x = median(as.numeric(d$.imp)),
        y = 0,
        label = "No \n imputations \n to show"
      ),
      color = "grey")
  }
  return(p)
}
#test
trace_one_variable(trace_dat, x = "hgt")
trace_one_variable(trace_dat, x = "age")
```


## Plot imputed data

We need to preprocess the `mids` object to make it `tidy`. Then we can plot the imputed data using one of four plotting devices:

- stripplot

- bwplot/boxplot

- densityplot

- xyplot/scatterplot


```{r}
# extract imputations and initialize plot
plot_imps <- function(imp, x, y = NULL) {
  # parse inputs
  if (is.null(y)) {
    y <- x
  }
  # combine observed and imputed data
  xy_obs <- imp$data %>%
    cbind(datapoint = "observed",
          .imp = 0,
          .id = 1:nrow(.),
          .) %>%
    .[!is.na(imp$data[[x]]) &
        !is.na(imp$data[[y]]), ]
  xy_imps <- imp %>% 
    mice::complete("long") %>% 
    cbind(datapoint = "imputed", .) %>% 
    .[.$.id %nin% xy_obs$.id,]
  xy_dat <- rbind(xy_obs, xy_imps) %>% 
    dplyr::mutate(datapoint = factor(datapoint, levels = c("observed", "imputed")))
  # initialize plot
  p <- xy_dat %>% 
    ggplot2::ggplot() +
    theme_mice() 
  # output
  return(p)
}

# test with continuous variable
x = "hc"

# boxplot (doesn't work for categorical variables)
imp %>% plot_imps(x) +
  ggplot2::geom_boxplot(
      ggplot2::aes(
        x = as.factor(.imp), 
        y = .data[[x]],
        color = datapoint
      ),
      width = 0.5
    ) +
  ggplot2::xlab("Imputation (0 = observed data)") 

# stripplot
imp %>% plot_imps(x) +
  ggplot2::geom_jitter(
      ggplot2::aes(
        x = as.factor(.imp),
        y = .data[[x]],
        color = datapoint
      ),
      height = 0.25,
      width = 0.25) +
    ggplot2::xlab("Imputation (0 = observed data)") 

# density plot
imp %>% plot_imps(x) +
  ggplot2::geom_density(ggplot2::aes(
    x = .data[[x]],
    color = datapoint,
    size = datapoint,
    group = .imp
  )) 

# test with two continuous variables
y = "hgt"

# xyplot
imp %>% plot_imps(x, y) +
  ggplot2::geom_point(ggplot2::aes(
    x = .data[[x]],
    y = .data[[y]],
    group = .id,
    color = datapoint
  ))

# test with categorical variable
x = "phb"

# faceted barplot
imp %>% plot_imps(x) +
  ggplot2::geom_bar(
      ggplot2::aes(
        x = .data[[x]],
        color = datapoint,
        size = datapoint
      ),
      width = 0.5,
      fill = "white"#,
      #position = ggplot2::position_dodge(preserve = "single")
    ) +
  ggplot2::facet_wrap(~.imp, scales = "free_y", ncol = 2)

```

Maybe develop a predictor matrix to adjust for the imputation model?

```{r fig.height=6, fig.width=6}
# plot the predictor matrix for the imputation model, make interactive later!
plot_pred_matrix <- function(dat) {
  # parse input
  if (mice::is.mids(dat)) {
    imp <- dat
  } else {
    imp <- mice::mice(dat, maxit = 0)
  }
  # get the predictor matrix and add the rownames as variable
  pred <- imp$predictorMatrix %>%
    as.data.frame() %>%
    cbind(vrb_to_imp = rownames(.), .)
  # make the data tidy to plot
  long_pred <-
    pivot_longer(
      pred,
      cols = names(pred)[-1],
      names_to = "vrb_as_pred",
      names_transform = list(vrb_as_pred = as.factor)
    )
  # plot the predictor matrix
  p <- long_pred %>%
    ggplot2::ggplot() +
    ggplot2::geom_tile(ggplot2::aes(
      x = vrb_as_pred,
      y = vrb_to_imp,
      fill = as.factor(value)
    ),
    color = "black") +
    ggplot2::theme(legend.position = "none", panel.grid = ggplot2::element_blank()) +
    ggplot2::scale_y_discrete(limits = rev(names(imp$data))) +
    ggplot2::scale_x_discrete(limits = names(imp$data), position = "top") +
    ggplot2::scale_fill_manual(values = c("1" = mice:::mdc(1), "0" = mice:::mdc(2))) +
    ggplot2::labs(x = "Predictor", y = "Variable to impute", title = "Predictor matrix")
  
  return(p)
}

# test
plot_pred_matrix(imp)
new_pred <- imp$predictorMatrix
new_pred["bmi", c("hgt", "wgt")] <- 0
imp2 <- mice::mice(dat, predictorMatrix = new_pred, maxit = 0)
plot_pred_matrix(imp2)

```

