---
title: "ggmice"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggmice}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# set-up environment
library(shinymice)
# example dataset
dat <- mice::boys
# example mids object
imp <- mice::mice(dat, printFlag = FALSE)
```

## Plot incomplete data

We can evaluate the missingness in the incomplete data using two plotting devices:

- histogram

- scatterplot


```{r}
# histogram plot

```

```{r}
# scatterplot

```

## Plot imputed data

We need to preprocess the `mids` object to make it `tidy`. Then we can plot the imputed data using one of four plotting devices:

- stripplot

- boxplot

- densityplot

- scatterplot

```{r}
# preprocess
prepare_plotting <- function(mids, x, y = NULL) {
  imps <- mids$imp[[x]] %>%
    dplyr::mutate(.id = row.names(.)) %>%
    tidyr::pivot_longer(
      cols = 1:mids$m,
      names_to = ".imp",
      values_to = x,
      names_transform = list(.imp = as.numeric)
    )
  
  # preprocess incomplete data
  d <-
    mids$data %>%
    dplyr::mutate(.id = rownames(.)) %>%
    .[c(x, y, ".id")] %>%
    cbind(.imp = 0)
  
  # combine incomplete and imputed data
  if (is.null(y)) {
    cd <- d %>%
      dplyr::bind_rows(., imps) %>%
      dplyr::mutate(imputed = ifelse(.imp > 0, 'Imputed', 'Observed'))
  }
  return(cd)
}

# test
x = "hc"
y = NULL
plot_dat <- prepare_plotting(mids = imp, x)
```

Now plot the prepared data

```{r}
# stripplot
m = 5
geom_stripplot <- function(x, m) {
  p <- list(
    ggplot2::geom_jitter(
      ggplot2::aes(
        x = as.factor(.data$.imp),
        y = .data[[x]],
        color = .data$imputed
      ),
      na.rm = TRUE,
      height = 0.1,
      width = 0.1
    ),
    ggplot2::scale_x_discrete(limits = as.character(0:m)),
    ggplot2::labs(
      x = "Imputation (0 = observed data)",
      y = paste0("Variable: '", x, "'"),
      color = "",
      fill = "",
      size = ""
    )
  )
  return(p)
}

plot_dat %>%
  ggplot2::ggplot() +
  geom_stripplot(x, m)
```

Now add a mice theme

```{r}
theme_mice <- function() {
  ggplot2::update_geom_defaults("point", list(
    shape = 21,
    size = 1.9,
    stroke = 1.1,
    alpha = 0.5
  ))
  
  theme <- list(
    ggplot2::theme_classic(),
    ggplot2::theme(legend.position = "bottom"),
    ggplot2::scale_color_manual(
      values = c(
        "Observed" = mice:::mdc(1),
        "Missing" = mice:::mdc(2),
        "Imputed" = mice:::mdc(2)
      )
    ),
    ggplot2::scale_fill_manual(
      values = c(
        "Observed" = mice:::mdc(1),
        "Missing" = mice:::mdc(2),
        "Imputed" = mice:::mdc(2)
      )
    )
  )
  return(theme)
}

plot_dat %>%
  ggplot2::ggplot() +
  geom_stripplot(x, m) +
  theme_mice()
```

Do the same for a boxplot

```{r}
geom_boxplot <- function(x, m){
  list(
    ggplot2::stat_boxplot(
      ggplot2::aes(
        x = as.factor(.data$.imp),
        y = .data[[x]],
        color = .data$imputed
      ),
      na.rm = TRUE,
      width = 0.25,
      size = 1,
      geom = "errorbar"
    ),
    ggplot2::geom_boxplot(
      ggplot2::aes(
        x = as.factor(.data$.imp),
        y = .data[[x]],
        color = .data$imputed
      ),
      na.rm = TRUE,
      size = 1
    ),
    ggplot2::scale_x_discrete(limits = as.character(0:m)),
    ggplot2::labs(
      x = "Imputation (0 = observed data)",
      y = paste0("Variable: '", x, "'"),
      color = "",
      fill = "",
      size = ""
    ))
}

plot_dat %>% 
  ggplot2::ggplot() +
  geom_boxplot(x, m) + 
  theme_mice()

```

