---
title: "ggmice"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggmice}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE, 
  message = FALSE,
  fig.width = 6, 
  fig.height = 6
)
```

```{r setup, include = FALSE}
# set-up environment
# library(shinymice)
library(tidyverse)
`%nin%` <- Negate(`%in%`)

# example dataset
dat <- mice::boys
# example mids object
imp <- mice::mice(dat, printFlag = FALSE)

# define mice theme
theme_mice <- function() {
  # move this outside of the function to unnecessary avoid re-running
  ggplot2::update_geom_defaults("point", list(
    shape = 21,
    size = 1.9,
    stroke = 1.1,
    alpha = 0.5
  ))
  ggplot2::update_geom_defaults("boxplot", list(
    size = 1
  ))
  ggplot2::update_geom_defaults("line", list(
    size = 1
  ))
  # the actual theme settings
  theme <- list(
    ggplot2::theme_classic(),
    ggplot2::theme(legend.position = "top"),
    ggplot2::scale_color_manual(
      values = c(
        "observed" = mice:::mdc(1),
        "missing" = mice:::mdc(2),
        "imputed" = mice:::mdc(2)
      )
    ),
    ggplot2::scale_fill_manual(
      values = c(
        "observed" = mice:::mdc(1),
        "missing" = mice:::mdc(2),
        "imputed" = mice:::mdc(2)
      )
    ),
    ggplot2::scale_size_manual(
      values = c("observed" = 1,
      "imputed" = 0.5)
),
    ggplot2::labs(color = NULL, fill = NULL, size = NULL)
  )
  return(theme)
}
```

## Plot incomplete data

We can evaluate the missingness in the incomplete data using two plotting devices:

- histogram

- scatterplot


```{r}
# histogram/density plot dodged instead of faceted
# plot_conditional <- function(dat, x, z) {
#   if (is.numeric(dat[[x]])) {
#     geom <- ggplot2::geom_density(ggplot2::aes(x = .data[[x]], y = ..count.., color = conditional))
#   } else {
#     geom <-
#       ggplot2::geom_bar(
#         ggplot2::aes(x = .data[[x]], color = conditional),
#         position = ggplot2::position_dodge(),
#         fill = "white"
#       )
#   }
#   dat %>%
#     dplyr::mutate(conditional = factor(
#       is.na(.data[[z]]),
#       levels = c(TRUE, FALSE),
#       labels = c("missing", "observed")
#     )) %>%
#     ggplot2::remove_missing(., vars = x) %>%
#     ggplot2::ggplot() +
#     geom +
#     theme_mice() +
#     ggplot2::labs(color = paste(z)) +
#     theme(legend.position = "top")
# }

# histogram/density plot faceted instead of dodged
plot_conditional <- function(dat, x, z) {
  # define graphing elements to add to plot
  if (is.numeric(dat[[x]])) { # for continuous variables
    geom <-
      ggplot2::geom_histogram()
  } else { # for categorical variables
    geom <-
      ggplot2::geom_bar()
  }
  
  # preprocess the data
  dat[!is.na(dat[[x]]), ] %>%
    dplyr::mutate(conditional = factor(
      is.na(.data[[z]]),
      levels = c(FALSE, TRUE),
      labels = c("observed", "missing")
    )) %>%
    # plot
    ggplot2::ggplot(ggplot2::aes(
        x = .data[[x]],
        color = conditional,
        fill = "white"
      )) +
    geom +
    # split by conditional variable
    ggplot2::facet_wrap(
      ~ conditional,
      ncol = 1,
      scales = "free_y",
      labeller = ggplot2::labeller(conditional = c(
        paste(z, "observed"), paste(z, "missing")
      ) %>% setNames(c(
        "observed", "missing"
      )))
    ) +
    # style
    theme_mice() +
    theme(legend.position = "none",
          strip.background = element_rect(size = 0.5))
}

# test
plot_conditional(dat, x = "age", z = "hc")
plot_conditional(dat, x = "reg", z = "hc")
```


```{r}
# scatterplot with NAs
plot_bivariate <- function(dat, x, y) {
  # define graphing elements to add to plot later
  # set NA value and scale for variable x
  if (is.numeric(dat[[x]])) {
    NA_x <-
      min(dat[[x]], na.rm = TRUE) - .1 * diff(range(dat[[x]], na.rm = TRUE))
    shade_x <-
      min(dat[[x]], na.rm = TRUE) - .05 * diff(range(dat[[x]], na.rm = TRUE))
    scale_x <- NULL
  } else {
    NA_x <- "NA"
    shade_x <- 1.5
    scale_x <-
      ggplot2::scale_x_discrete(limits = c("NA", levels(dat[[x]])))
  }
  
  # set NA value and scale for variable y
  if (is.numeric(dat[[y]])) {
    NA_y <-
      min(dat[[y]], na.rm = TRUE) - .1 * diff(range(dat[[y]], na.rm = TRUE))
    shade_y <-
      min(dat[[y]], na.rm = TRUE) - .05 * diff(range(dat[[y]], na.rm = TRUE))
    scale_y <- NULL
  } else {
    NA_y <- "NA"
    shade_y <- 1.5
    scale_y <-
      ggplot2::scale_y_discrete(limits = c("NA", levels(dat[[y]])))
  }
  
  # add geom for NAs in the x space
  if (any(is.na(dat[[x]]))) {
    geom_x <- ggplot2::geom_point(
      position = ggplot2::position_jitter(width = 0.2, height = 0.2),
      color = mice:::mdc(2),
      shape = 4,
      mapping = ggplot2::aes(x = NA_x, y = .data[[y]]),
      data = dat[is.na(dat[[x]]),]
    )
  } else {
    geom_x <- NULL
  }
  
  # add geom for NAs in the y space
  if (any(is.na(dat[[y]]))) {
    geom_y <- ggplot2::geom_point(
      position = ggplot2::position_jitter(width = 0.2, height = 0.2),
      color = mice:::mdc(2),
      shape = 4,
      mapping = ggplot2::aes(x = .data[[x]], y = NA_y),
      data = dat[is.na(dat[[y]]),]
    )
  } else{
    geom_y <- NULL
  }
  
  # add geom for NAs in the x-y space
  if (any(is.na(dat[[x]]) & is.na(dat[[y]]))) {
    geom_xy <- ggplot2::geom_point(
      position = ggplot2::position_jitter(width = 0.2, height = 0.2),
      color = mice:::mdc(2),
      shape = 4,
      mapping = ggplot2::aes(x = NA_x, y = NA_y),
      data = dat[is.na(dat[[x]]) & is.na(dat[[y]]),]
    )
  } else{
    geom_xy <- NULL
  }
  
  # plot
  p <- dat %>% ggplot2::ggplot() +
    # plot observed datapoints in x-y space
    ggplot2::geom_point(
      ggplot2::aes(x = .data[[x]], y = .data[[y]]),
      position = ggplot2::position_jitter(width = 0.2, height = 0.2),
      color = mice::mdc(1)
    ) +
    # add area for NAs in the x space
    ggplot2::annotate(
      "rect",
      xmin = -Inf,
      xmax = shade_x,
      ymin = -Inf,
      ymax = Inf,
      alpha = 0.1
    ) +
    # add area for NAs in the y space
    ggplot2::annotate(
      "rect",
      xmin = -Inf,
      xmax = Inf,
      ymin = -Inf,
      ymax = shade_y,
      alpha = 0.1
    ) +
    # add NAs 
    geom_x +
    geom_y +
    geom_xy +
    # style graph
    theme_mice() +
    scale_x +
    scale_y 
  
  # output
  return(p)
}

# test
plot_bivariate(dat, x = "reg", y = "hc")
plot_bivariate(dat, x = "hgt", y = "hc") 
```

## Plot missingness

- md pattern

- fluxplot

First look at the missing data pattern

```{r fig.height=7.5, fig.width=6}
# make ggplot version of md pattern plot
plot_md_pattern <- function(dat) {
  # get md pattern and store additional info
  pat <- mice::md.pattern(dat)
  vrb <- colnames(pat)[-ncol(pat)]
  colnames(pat) <- c(vrb, "n_vrb_inc")
  n_pat_obs <- as.numeric(rownames(pat))[-nrow(pat)]
  n_vrb_inc <- as.numeric(pat[, ncol(pat)])
  n_val_mis <- as.numeric(pat[nrow(pat), ])[-ncol(pat)]
  
  # make the pattern tidy
  long_pat <- pat[-nrow(pat), ] %>%
    cbind(., n_pat_obs, pat_nr = 1:nrow(.)) %>%
    as.data.frame() %>%
    tidyr::pivot_longer(cols = all_of(vrb),
                        names_to = "vrb",
                        values_to = "obs") %>%
    cbind(., n_val_mis) %>%
    dplyr::mutate(pat_nr = factor(pat_nr, levels = 1:max(pat_nr)))
  
  # plot the md pattern
  p <- long_pat %>%
    ggplot2::ggplot() +
    ggplot2::geom_tile(ggplot2::aes(
      x = vrb,
      y = pat_nr,
      fill = as.factor(obs),
      group = n_vrb_inc
    ),
    color = "black") +
    ggplot2::theme(legend.position = "none") +
    ggplot2::scale_fill_manual(values = c("1" = mice:::mdc(1), "0" = mice:::mdc(2))) +
    ggplot2::scale_y_discrete(limits = rev(levels(long_pat$pat_nr)),
                              labels = rev(as.character(n_pat_obs))) +
    ggplot2::scale_x_discrete(limits = vrb, position = "bottom", labels = as.character(n_val_mis)) + #, labels = as.character(n_val_mis)) +
    ggplot2::labs(
      x = NULL, 
      y = "frequency"
      ) + #, 
      #title = "Missing data pattern (patterns as rows, variables as columns)") +
    geom_text(aes(x = vrb, y = Inf, label = vrb), data = long_pat, vjust = -0.5) +
    coord_cartesian(clip = "off") +
    theme(plot.margin = margin(20, 2, 2, 2, unit = "pt"))

  
  return(p)
}

# test
plot_md_pattern(dat)
```

Then look at the in- and outflux

```{r}
# plot in- and outflux of incomplete data
plot_flux <- function(dat){
  flx <- mice::flux(dat)
  p <- flx %>% ggplot2::ggplot() +
    ggplot2::geom_text(
      ggplot2::aes(x = influx, y = outflux, label = rownames(flx))
    ) +
    geom_abline(intercept = 1, slope = -1, linetype = "dashed") +
    lims(x = c(0,1), y = c(0,1)) +
    theme_classic()
  return(p)
}
# test
plot_flux(dat)
```


## Plot convergence

```{r}

# make chain means and variances tidy
preprocess_thetas <- function(imp) {
  d <- imp$chainMean %>%
    dplyr::na_if(., "NaN") %>%
    as.data.frame() %>%
    dplyr::mutate(var = row.names(.), theta = "Chain means") %>%
    rbind(
      .,
      imp$chainVar %>%
        sqrt() %>%
        as.data.frame(.) %>%
        dplyr::mutate(var = row.names(.), theta = "Chain standard deviations")
    ) %>%
    tidyr::pivot_longer(-c(var, theta)) %>%
    cbind(.it = as.integer(1:imp$iteration),
          .imp = as.factor(rep(1:imp$m, each = imp$iteration)))
  return(d)
}
# test
trace_dat <- preprocess_thetas(imp)

trace_one_variable <- function(d, x) {
  # plot one variable
  p <- d[d$var == x,] %>%
    ggplot2::ggplot() +
    ggplot2::geom_line(ggplot2::aes(x = .it, y = value, color = .imp)) +
    ggplot2::facet_wrap( ~ theta, scales = "free", ncol = 2) +
    ggplot2::theme_classic() +
    ggplot2::theme(
      strip.background = element_rect(colour = "white"),
      strip.text = element_text(face = "bold", size = 11),
      legend.position = "bottom"
    ) +
    ggplot2::labs(x = "Iteration",
                  y = paste0("Variable: '", x, "'"),
                  color = "Imputation")
  # show message to user when variable is completely observed
  if (all(is.na(p$data$value))) {
    p <-
      p + ggplot2::geom_text(ggplot2::aes(
        x = median(as.numeric(d$.imp)),
        y = 0,
        label = "No \n imputations \n to show"
      ),
      color = "grey")
  }
  return(p)
}
#test
trace_one_variable(trace_dat, x = "hgt")
trace_one_variable(trace_dat, x = "age")
```


## Plot imputed data

We need to preprocess the `mids` object to make it `tidy`. Then we can plot the imputed data using one of four plotting devices:

- stripplot

- bwplot/boxplot

- densityplot

- xyplot/scatterplot

```{r}
# preprocess
prepare_plotting <- function(mids, x, y = NULL) {
  # preprocess incomplete data
  d <-
    mids$data %>%
    dplyr::mutate(.id = rownames(.)) %>%
    .[c(x, y, ".id")] %>%
    cbind(.imp = 0)
  
  # preprocess imputations
  imps <- mids$imp[[x]] %>%
    dplyr::mutate(.id = row.names(.)) %>%
    tidyr::pivot_longer(
      cols = 1:mids$m,
      names_to = ".imp",
      values_to = x,
      names_transform = list(.imp = as.numeric)
    ) 
  
  # combine incomplete and imputed data
    cd <- d %>%
      dplyr::bind_rows(., imps) %>%
      dplyr::mutate(imputed = ifelse(.imp > 0, 'imputed', 'observed'))
  
  # output
  return(cd)
}

# test 
x = "hc"
y = "reg"
plot_dat <- prepare_plotting(mids = imp, x = x, y = y)
# # with second variable
# plot_dat2 <- prepare_plotting(mids = imp, x = y, y = x)
# a <- dplyr::full_join(plot_dat, plot_dat2)
```

Now plot the prepared data

```{r}
# stripplot
m = 5
geom_stripplot <- function(x, m, ...) {
  p <- list(
    ggplot2::geom_jitter(
      ggplot2::aes(
        x = as.factor(.data$.imp),
        y = .data[[x]],
        color = .data$imputed
      ),
      height = 0,
      width = 0.25, 
      ...
    ),
    ggplot2::scale_x_discrete(limits = as.character(0:m)),
    ggplot2::labs(
      x = "Imputation (0 = observed data)",
      y = paste0("Variable: '", x, "'"),
      color = "",
      fill = "",
      size = ""
    )
  )
  return(p)
}

# test
# plot_dat %>%
#   ggplot2::ggplot() +
#   geom_stripplot(x, m)
```

Now add a mice theme

```{r}

plot_dat %>%
  ggplot2::ggplot() +
  geom_stripplot(x, m) +
  theme_mice()
```

Do the same for a boxplot

```{r}
geom_bwplot <- function(x, m){
  list(
    ggplot2::geom_boxplot(
      ggplot2::aes(
        x = as.factor(.data$.imp),
        y = .data[[x]],
        color = .data$imputed
      ),
      width = 0.5
    ),
    ggplot2::scale_x_discrete(limits = as.character(0:m)),
    ggplot2::xlab(
      "Imputation (0 = observed data)")
    )
}

plot_dat %>% 
  ggplot2::ggplot() +
  geom_bwplot(x, m = imp$m) + 
  theme_mice()

# try to combine this with the stripplot??
plot_dat %>%
  ggplot2::ggplot() +
  geom_bwplot(x, m) + 
  geom_stripplot(x, m, alpha = 0.1) +
  theme_mice()

```

And something similar for density plots:

```{r}
geom_densityplot <- function(x, m){
  list(ggplot2::geom_density(
    ggplot2::aes(
    x = .data[[x]],
    group = .data$.imp,
    color = .data$imputed,
    size = .data$imputed)),
    ggplot2::scale_size_manual(guide = "none", values = c(0.5, 1) %>% setNames(c("imputed", "observed")))
)}


plot_dat[!is.na(plot_dat[[x]]), ] %>%  
  ggplot2::ggplot() +
  geom_densityplot(x, m = imp$m) + 
  theme_mice()


```
<!-- Note the addition of `na.omit()` to make sure that categorical variables don't get an NA category. -->

```{r}
# At some point, try to do this for histograms too? See https://github.com/amices/shinyMice/blob/Enhancements/R/gg.barplot.R

# geom_densityplot <- function(x, m){
#   list(ggplot2::geom_density(
#     ggplot2::aes(
#     x = .data[[x]],
#     group = .data$.imp,
#     color = .data$imputed,
#     size = .data$imputed)),
#     ggplot2::scale_size_manual(values = c(0.5, 1) %>% setNames(c("imputed", "observed")))
# )}
# 
# plot_dat %>% 
#   ggplot2::ggplot() +
#   geom_densityplot(x, m = imp$m) + 
#   theme_mice()
```

Now for something different: xyplots!

```{r}
# better preprocessing function??
x = "hgt"
y = "hc"

# extract imputations
imps <- imp %>%
  mice::complete("long")

preprocess_imps <- function(imp, x, y = NULL) {
  if (is.null(y)) {
    y <- x
  }
  # combine observed and imputed data
  xy_obs <- imp$data %>%
    cbind(datapoint = "observed",
          .imp = 0,
          .id = 1:nrow(.),
          .) %>%
    .[!is.na(imp$data[[x]]) &
        !is.na(imp$data[[y]]), ]
  xy_imps <- imps %>% 
    cbind(datapoint = "imputed", .) %>% 
    .[imps$.id %nin% xy_obs$.id,]
  xy_dat <- rbind(xy_obs, xy_imps) %>% 
    dplyr::mutate(datapoint = factor(datapoint, levels = c("observed", "imputed")))
  return(xy_dat)
}


xy_dat <- imp %>% preprocess_imps(x = "hgt", y = "hc")
# two implementations, last one is more efficient
# xy_obs %>%
#   ggplot2::ggplot(ggplot2::aes(
#     x = .data[[x]],
#     y = .data[[y]],
#     group = .id,
#     color = datapoint
#   )) +
#   ggplot2::geom_point() +
#   ggplot2::geom_point(data = xy_imps)  +
#   theme_mice() +
#   labs(color = NULL)

xy_dat %>%
  ggplot2::ggplot() +
  ggplot2::geom_point(ggplot2::aes(
    x = .data[[x]],
    y = .data[[y]],
    group = .id,
    color = datapoint
  )) +
  theme_mice() +
  labs(color = NULL)

# geom_scatter <- ggplot2::geom_point(ggplot2::aes(
#     x = .data[[x]],
#     y = .data[[y]],
#     group = .id,
#     color = datapoint
#   ))

# maybe do something like this for the other imp plots??
x = "hc"
imp %>% preprocess_imps(x) %>% 
  ggplot2::ggplot() +
  theme_mice() +
  ggplot2::geom_boxplot(
      ggplot2::aes(
        x = as.factor(.data$.imp), #x = factor(.data$.imp, levels = c(0, 1:imp$m), ordered = TRUE), use icw   scale_x_discrete(drop=FALSE)
        y = .data[[x]],
        color = .data$datapoint
      ),
      width = 0.5
    ) +
  ggplot2::geom_jitter(
      ggplot2::aes(
        x = as.factor(.imp),
        y = .data[[x]],
        color = datapoint
      ),
      height = 0,
      width = 0.25,
      alpha = 0.1) +
  ggplot2::xlab("Imputation (0 = observed data)") 

# density plot
imp %>% preprocess_imps(x) %>%
  ggplot2::ggplot() +
  theme_mice() +
  ggplot2::geom_density(ggplot2::aes(
    x = .data[[x]],
    color = datapoint,
    size = datapoint,
    group = .imp
  )) 

# bar plot for categorical variables
imp %>% preprocess_imps(x) %>%
  ggplot2::ggplot() +
  theme_mice() +
  ggplot2::geom_histogram(ggplot2::aes(
    x = .data[[x]],
    color = datapoint,
    group = .imp
  ),
  fill = "white",
  position = "stack") +
  ggplot2::facet_wrap(~datapoint, scales = "free_y", ncol = 1) +
  ggplot2::theme(legend.position = "none")


x = "reg"
imp %>% preprocess_imps(x) %>%
  ggplot2::ggplot() +
  theme_mice() +
  ggplot2::geom_bar(
    ggplot2::aes(x = .data[[x]],
                 color = datapoint,
                 group = .imp),
    fill = "white",
    position = "stack"
  ) +
  ggplot2::facet_wrap(
    ~ datapoint,
    scales = "free_y",
    ncol = 1,
    labeller = ggplot2::labeller(datapoint = c(paste(x, "observed"), paste(x, "imputed")) %>% 
                                   setNames(c("observed", "imputed"))
                                 )
  ) +
  ggplot2::theme(legend.position = "none")

```

Maybe develop a predictor matrix to adjust for the imputation model?

```{r fig.height=6, fig.width=6}
# plot the predictor matrix for the imputation model, make interactive later!
plot_pred_matrix <- function(dat) {
  # parse input
  if (mice::is.mids(dat)) {
    imp <- dat
  } else {
    imp <- mice::mice(dat, maxit = 0)
  }
  # get the predictor matrix and add the rownames as variable
  pred <- imp$predictorMatrix %>%
    as.data.frame() %>%
    cbind(vrb_to_imp = rownames(.), .)
  # make the data tidy to plot
  long_pred <-
    pivot_longer(
      pred,
      cols = names(pred)[-1],
      names_to = "vrb_as_pred",
      names_transform = list(vrb_as_pred = as.factor)
    )
  # plot the predictor matrix
  p <- long_pred %>%
    ggplot2::ggplot() +
    ggplot2::geom_tile(ggplot2::aes(
      x = vrb_as_pred,
      y = vrb_to_imp,
      fill = as.factor(value)
    ),
    color = "black") +
    ggplot2::theme(legend.position = "none") +
    ggplot2::scale_y_discrete(limits = rev(names(imp$data))) +
    ggplot2::scale_x_discrete(limits = names(imp$data), position = "top") +
    ggplot2::scale_fill_manual(values = c("1" = mice:::mdc(1), "0" = mice:::mdc(2))) +
    ggplot2::labs(x = "Predictor", y = "Variable to impute", title = "Predictor matrix")
  
  return(p)
}

# test
plot_pred_matrix(imp)
new_pred <- imp$predictorMatrix
new_pred["bmi", c("hgt", "wgt")] <- 0
imp2 <- mice::mice(dat, predictorMatrix = new_pred, maxit = 0)
plot_pred_matrix(imp2)

```

