---
title: "ggmice"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggmice}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE, 
  message = FALSE,
  fig.width = 6, 
  fig.height = 6
)
```

```{r setup, include = FALSE}
# set-up environment
library(shinymice)
library(tidyverse)

# example dataset
dat <- mice::boys
# example mids object
imp <- mice::mice(dat, printFlag = FALSE)
```

## Plot incomplete data

We can evaluate the missingness in the incomplete data using two plotting devices:

- histogram

- scatterplot


```{r}
# conditional distribution
shinymice::plot_NA_cond(dat, x = "age", z = "hc")
shinymice::plot_NA_cond(dat, x = "reg", z = "hc")
```


```{r}
# scatterplot with NAs
shinymice::plot_NA_scatter(dat, x = "reg", y = "hc")
shinymice::plot_NA_scatter(dat, x = "hgt", y = "hc")
```


```{r eval=FALSE, include=FALSE}
# # same same but easier!
# x = "hgt"
# y = "hc"
# dat %>% ggplot2::ggplot() +
#   ggplot2::theme_classic() +
#   # plot observed datapoints in x-y space
#   ggplot2::geom_point(
#     ggplot2::aes(x = .data[[x]], y = .data[[y]]),
#     position = ggplot2::position_jitter(width = 0.2, height = 0.2),
#     color = mice::mdc(1)
#   ) +
#   # missing x
#   ggplot2::geom_point(
#     position = ggplot2::position_jitter(width = 0.2, height = 0.2),
#     color = mice:::mdc(2),
#     shape = 4,
#     mapping = ggplot2::aes(x = -Inf, y = .data[[y]]),
#     data = dat[is.na(dat[[x]]),]
#   ) +
#   # missing y
#   ggplot2::geom_point(
#     position = ggplot2::position_jitter(width = 0.2, height = 0.2),
#     color = mice:::mdc(2),
#     shape = 4,
#     mapping = ggplot2::aes(x = .data[[x]], y = -Inf),
#     data = dat[is.na(dat[[y]]),]
#   ) +
#   # missing xy
#   ggplot2::geom_point(
#     position = ggplot2::position_jitter(width = 0.2, height = 0.2),
#     color = mice:::mdc(2),
#     shape = 4,
#     mapping = ggplot2::aes(x = -Inf, y = -Inf),
#     data = dat[is.na(dat[[x]]) & is.na(dat[[y]]),]
#   ) +
#   # axes
#   ggplot2::coord_cartesian(clip = "off") 

```

## Plot imputation model/missingness

- md pattern

- fluxplot

First look at the missing data pattern

```{r fig.height=7.5, fig.width=6}
# plot md pattern 
dat2 <- mice::boys %>% cbind(bhgdhjahfjhksgdkfkjhsdgfksd = mice::boys$age) 
shinymice::plot_md_pattern(dat2) 
pat <- mice::md.pattern(dat)
```

Then look at the in- and outflux

```{r}
# plot in- and outflux of incomplete data
shinymice::plot_flux(dat)
```
Develop a predictor matrix to adjust for the imputation model

```{r fig.height=6, fig.width=6}
# plot the predictor matrix for the imputation model
shinymice::plot_pred_matrix(imp)
# change the predictor matrix and plot again
new_pred <- imp$predictorMatrix
new_pred["bmi", c("hgt", "wgt")] <- 0
imp2 <- mice::mice(dat, predictorMatrix = new_pred, maxit = 0)
shinymice::plot_pred_matrix(imp2)
```


## Plot convergence

```{r}
# traceplot: make chain means and variances tidy
trace_dat <- shinymice::preprocess_thetas(imp)

# traceplot: plot trace of one variable
shinymice::trace_one_variable(trace_dat, x = "hgt")
shinymice::trace_one_variable(trace_dat, x = "age")
```


## Plot imputed data

We need to preprocess the `mids` object to make it `tidy`. Then we can plot the imputed data using one of four plotting devices:

- stripplot

- bwplot/boxplot

- densityplot

- xyplot/scatterplot


```{r}
# test imputation plots with continuous variable
x = "hc"

# boxplot 
shinymice::plot_bw(imp, x)

# stripplot
shinymice::plot_strip(imp, x)

# density plot
shinymice::plot_dens(imp, x)

# test with two continuous variables
y = "hgt"

# xyplot
shinymice::plot_xy(imp, x, y)

# test with categorical variable
x = "phb"

# boxplot (not informative with categorical variable)
shinymice::plot_bw(imp, x)

# stripplot
shinymice::plot_strip(imp, x)

# density plot (not appropriate with categorical variable)
shinymice::plot_dens(imp, x)

facet_labs <- c("Observed data", paste("Imputation", 1:5)) %>% setNames(0:5)
# faceted barplot
imp %>% plot_imps(x) +
  ggplot2::geom_bar(
    ggplot2::aes(x = .data[[x]],
                 color = datapoint),
    width = 0.5,
    fill = "white"#,
    #position = ggplot2::position_dodge(preserve = "single")
  ) +
  ggplot2::facet_wrap(~ .imp, scales = "free_y", ncol = 2, labeller = ggplot2::labeller(.imp = facet_labs))
# add arg 'showall = FALSE' 

```


## Additional: compute rhat

```{r}
# compute rhat
get.rhat <- function(sims) {
  # compute potential scale reduction factor (rhat) for each variable in mids object
  # equations adapted from Vehtari et al. (2019)
  # parse inputs
  n_it <- length(sims)
  # helper functions
  # split chains with maxit > 4 to detect trending
  split_chains <- function(sims) {
    # split Markov chains, adapted from rstan
    n_it <- dim(sims)[1]
    # output
    if (n_it < 4)
      # do not split if result will be chains of length 1
      return(sims)
    else {
      # split each chain to get 2m chains
      lower <- 1:floor(n_it / 2)
      upper <- ceiling((n_it / 2) + 1):n_it
      splits <- base::cbind(sims[lower,], sims[upper,])
      return(splits)
    }
  }
  # rank-normalize chains because Gelman says so
  z_scale <- function(x) {
    # rank-normalize Markov chain, copied from rstan
    t <- length(x)
    r <- rank(x, ties.method = 'average')
    z <- qnorm((r - 1 / 2) / t)
    # output
    if (!is.null(dim(x))) {
      # output should have the input dimensions
      z <- array(z, dim = dim(x), dimnames = dimnames(x))
    }
    return(z)
  }
  # preprocess chains
  sims <- sims %>% split_chains(.) %>% z_scale(.)
  # compute rhat
  var_between <-
    n_it * var(apply(sims, 2, mean))
  var_within <- mean(apply(sims, 2, var))
  rhat <-
    sqrt((var_between / var_within + n_it - 1) / n_it)
  # output
  return(rhat)
}

# function for Rhat for two or more imputation chains
compute_rhat <- function(x) {
  # input: object with theta values (rows are iterations, columns are imputations)
  # output: convergence diagnostic Rhat across imputations
  # parse inputs
  if (is.data.frame(x)) {
    x <- as.matrix(x)
  }
  # quit function if there are not enough iterations
  n_it <- dim(x)[1]
  if (is.null(n_it)) {
    return(data.frame(
      iteration = 1,
      max.r.hat = NA,
      r.hat = NA
    ))
  } else {
    r.hat <- map_dfr(2:n_it, function(it) {
      # compute r hat in all ways described by Vehtari et al. (2019)
      rhat_bulk <- x[1:it, ] %>%
        get.rhat(.)
      rhat_tail <- abs(x[1:it, ] - median(x[1:it, ])) %>% # fold sims
        get.rhat(.)
      max(rhat_bulk, rhat_tail) %>%
        data.frame(rhat = .)
    }) %>%
      rbind(NA, .) %>%
      cbind(iteration = 1:n_it, .)
  }
  # output
  return(r.hat)
}
# imp2 <- mice::mice.mids(imp, maxit = 15)
x = "hc"
imp3 <- mice::mice.mids(imp, maxit = 10, printFlag = FALSE) 
theta <- imp3$chainMean[x,,]
compute_rhat(theta) %>% 
  ggplot2::ggplot() + 
  ggplot2::geom_line(aes(x = iteration, y = rhat)) + 
  ggplot2::geom_hline(yintercept = 1.2, color = "grey", size = 1) + 
  ggplot2::theme_classic()

```

