---
title: "ggmice"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggmice}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# set-up environment
library(shinymice)
library(tidyverse)
# example dataset
dat <- mice::boys
# example mids object
imp <- mice::mice(dat, printFlag = FALSE)

# define mice theme
theme_mice <- function() {
  # move this outside of the function to unnecessary avoid re-running
  ggplot2::update_geom_defaults("point", list(
    shape = 21,
    size = 1.9,
    stroke = 1.1,
    alpha = 0.5
  ))
  ggplot2::update_geom_defaults("boxplot", list(
    size = 1
  ))
  ggplot2::update_geom_defaults("line", list(
    size = 1
  ))
  # the actual theme settings
  theme <- list(
    ggplot2::theme_classic(),
    ggplot2::theme(legend.position = "bottom"),
    ggplot2::scale_color_manual(
      values = c(
        "Observed" = mice:::mdc(1),
        "Missing" = mice:::mdc(2),
        "Imputed" = mice:::mdc(2)
      )
    ),
    ggplot2::scale_fill_manual(
      values = c(
        "Observed" = mice:::mdc(1),
        "Missing" = mice:::mdc(2),
        "Imputed" = mice:::mdc(2)
      )
    )
  )
  return(theme)
}
```

## Plot incomplete data

We can evaluate the missingness in the incomplete data using two plotting devices:

- histogram

- scatterplot


```{r}
# histogram/density plot
plot_conditional <- function(dat, x, z) {
  if (is.numeric(dat[[x]])) {
    geom <- ggplot2::geom_density(ggplot2::aes(x = .data[[x]], y = ..count.., color = conditional))
  } else {
    geom <-
      ggplot2::geom_bar(
        ggplot2::aes(x = .data[[x]], color = conditional),
        position = ggplot2::position_dodge(),
        fill = "white"
      )
  }
  dat %>%
    dplyr::mutate(conditional = factor(
      is.na(.data[[z]]),
      levels = c(TRUE, FALSE),
      labels = c("Missing", "Observed")
    )) %>%
    ggplot2::remove_missing(., vars = x) %>%
    ggplot2::ggplot() +
    geom +
    theme_mice() +
    ggplot2::labs(color = paste0("Variable '", z, "' is:"))
}

plot_conditional(dat, x = "age", z = "hc")
plot_conditional(dat, x = "reg", z = "hc")
```

```{r}
# scatterplot
x = "wgt"
y = "hgt"
NA_y <- min(dat[[y]], na.rm = TRUE) - sd(dat[[y]], na.rm = TRUE)
NA_x <- min(dat[[x]], na.rm = TRUE) - sd(dat[[x]], na.rm = TRUE)

dat %>% ggplot2::ggplot() +
  ggplot2::geom_point(ggplot2::aes(x = .data[[x]], y = .data[[y]]), color = mice::mdc(1)) +
  ggplot2::geom_point(
    color = mice:::mdc(2),
    shape = 4,
    mapping = ggplot2::aes(x = .data[[x]], y = NA_y),
    data = dat[is.na(dat[[y]]),]
  ) +
  ggplot2::geom_point(
    color = mice:::mdc(2),
    shape = 4,
    mapping = ggplot2::aes(x = NA_x, y = .data[[y]]),
    data = dat[is.na(dat[[x]]),]
  ) +
  ggplot2::geom_point(
    color = mice:::mdc(2),
    shape = 4,
    mapping = ggplot2::aes(x = NA_x, y = NA_y),
    data = dat[is.na(dat[[x]]) &
                 is.na(dat[[y]]),]
  ) + ggplot2::theme_classic() +
  ggplot2::scale_x_continuous(expand = c(0.01, 0.01)) +
  ggplot2::scale_y_continuous(expand = c(0.01, 0.01)) 

```

# Plot missingness

- md pattern

- fluxplot

```{r}
# md pattern

```

```{r}
# plot in- and outflux of incomplete data
plot_flux <- function(dat){
  flx <- mice::flux(dat)
  p <- flx %>% ggplot2::ggplot() +
    ggplot2::geom_text(
      ggplot2::aes(x = influx, y = outflux, label = rownames(flx))
    ) +
    geom_abline(intercept = 1, slope = -1, linetype = "dashed") +
    lims(x = c(0,1), y = c(0,1)) +
    theme_classic()
  return(p)
}
plot_flux(flx)
```


## Plot imputed data

We need to preprocess the `mids` object to make it `tidy`. Then we can plot the imputed data using one of four plotting devices:

- stripplot

- bwplot/boxplot

- densityplot

- xyplot/scatterplot

```{r}
# preprocess
prepare_plotting <- function(mids, x, y = NULL) {
  imps <- mids$imp[[x]] %>%
    dplyr::mutate(.id = row.names(.)) %>%
    tidyr::pivot_longer(
      cols = 1:mids$m,
      names_to = ".imp",
      values_to = x,
      names_transform = list(.imp = as.numeric)
    )
  
  # preprocess incomplete data
  d <-
    mids$data %>%
    dplyr::mutate(.id = rownames(.)) %>%
    .[c(x, y, ".id")] %>%
    cbind(.imp = 0)
  
  # combine incomplete and imputed data
  if (is.null(y)) {
    cd <- d %>%
      dplyr::bind_rows(., imps) %>%
      dplyr::mutate(imputed = ifelse(.imp > 0, 'Imputed', 'Observed'))
  }
  return(cd)
}

# test
x = "hc"
y = NULL
plot_dat <- prepare_plotting(mids = imp, x)
```

Now plot the prepared data

```{r}
# stripplot
m = 5
geom_stripplot <- function(x, m) {
  p <- list(
    ggplot2::geom_jitter(
      ggplot2::aes(
        x = as.factor(.data$.imp),
        y = .data[[x]],
        color = .data$imputed
      ),
      height = 0.1,
      width = 0.1
    ),
    ggplot2::scale_x_discrete(limits = as.character(0:m)),
    ggplot2::labs(
      x = "Imputation (0 = observed data)",
      y = paste0("Variable: '", x, "'"),
      color = "",
      fill = "",
      size = ""
    )
  )
  return(p)
}

plot_dat %>%
  ggplot2::ggplot() +
  geom_stripplot(x, m)
```

Now add a mice theme

```{r}

plot_dat %>%
  ggplot2::ggplot() +
  geom_stripplot(x, m) +
  theme_mice()
```

Do the same for a boxplot

```{r}
geom_bwplot <- function(x, m){
  list(
    ggplot2::geom_boxplot(
      ggplot2::aes(
        x = as.factor(.data$.imp),
        y = .data[[x]],
        color = .data$imputed
      ),
    ),
    ggplot2::scale_x_discrete(limits = as.character(0:m)),
    ggplot2::labs(
      x = "Imputation (0 = observed data)",
      y = paste0("Variable: '", x, "'"),
      color = "",
      fill = "",
      size = ""
    ))
}

plot_dat %>% 
  ggplot2::ggplot() +
  geom_bwplot(x, m = imp$m) + 
  theme_mice()

```

And something similar for density plots:

```{r}
geom_densityplot <- function(x, m){
  list(ggplot2::geom_density(
    ggplot2::aes(
    x = .data[[x]],
    group = .data$.imp,
    color = .data$imputed,
    size = .data$imputed)),
    ggplot2::scale_size_manual(values = c(0.5, 1) %>% setNames(c("Imputed", "Observed")))
)}


plot_dat %>% na.omit() %>% 
  ggplot2::ggplot() +
  geom_densityplot(x, m = imp$m) + 
  theme_mice()


```
Note the addition of `na.omit()` to make sure that categorical variables don't get an NA category.

```{r}
# At some point, try to do this for histograms too? See https://github.com/amices/shinyMice/blob/Enhancements/R/gg.barplot.R

# geom_densityplot <- function(x, m){
#   list(ggplot2::geom_density(
#     ggplot2::aes(
#     x = .data[[x]],
#     group = .data$.imp,
#     color = .data$imputed,
#     size = .data$imputed)),
#     ggplot2::scale_size_manual(values = c(0.5, 1) %>% setNames(c("Imputed", "Observed")))
# )}
# 
# plot_dat %>% 
#   ggplot2::ggplot() +
#   geom_densityplot(x, m = imp$m) + 
#   theme_mice()
```

Now for something different: xyplots!

```{r}
geom_xyplot <- function(x, y, m) {
  list(ggplot2::geom_point(
    ggplot2::aes(
    x = .data[[x]],
    y = .data[[y]],
    color = .data$imputed
    )))}

plot_dat %>% na.omit() %>% 
  ggplot2::ggplot() +
  geom_densityplot(x, m = imp$m) + 
  theme_mice()
```

