---
title: "Convergence of MI algorithms"
output: html_notebook
---

This file describes how convergence diagnostic $\widehat{R}$ is computed in the simulation study I ran for my Research Report. $\widehat{R}$ is also called 'potential scale reduction factor'. It tells us how much the variance of an estimate could be shrunken down if we would have infinitely many iterations.

# Step 1: Impute some data

It's easiest to show how $\widehat{R}$ is computed by applying it on a simple example. Load the nhanes data and impute the variables with missingness.
```{r echo=TRUE, warning=FALSE}
# load required packages/functions
library(mice)
library(magrittr)
source("Functions/Convergence.R")
source("Functions/Convergence_supplement.R")

# set parameters
maxit <- 10
n.var <- 3 

# run imputation
imp <- mice(nhanes[,-1], maxit = maxit, print = F)

```

# Step 2: Evaluate the functions to use

We are going to compute $\widehat{R}$ values per imputed variable. As recommended by Vehtari et al. (2019), we compute split-half, rank-normalized $\widehat{R}$ values for the bulk and the tails of the distribution. For the latter, we need to fold the imputation chains before splitting them in half and rank-normalizing them. Then, $\widehat{R}$ is computed for both (bulk and tails). The highest of these two $\widehat{R}$ values is extracted. This is performed for each of the imputed variables, and only the maximum $\widehat{R}$ value across variables is reported. 

The rhat_function function is the wrapper that performs all of this. Within the function, we call 4 other functions (explaination below): fold_sims(), split_chains(), z_scale(), and get.rhat(). 

```{r}
# show the contents of the function
rhat_function

```


The first function inside rhat_function is split_chains(). This 'chops' the chains in two, to assess trending within chains. 
```{r warning=FALSE}
# show the contents of the function
split_chains

```

Then the halved chains are rank-normalized.
```{r}
# show the contents of the function
z_scale

```

To detect non-convergence in the tails, $\widehat{R}$ is also computed on the folded chains (which are also split and rank-normalized).
```{r}
# show the contents of the function
fold_sims

```

Only then do we get to the actual $\widehat{R}$ function. In the get.rhat function, the following equations are applied:

"In the equations below, $N$ is the number of draws per chain, $M$ is the number of chains, and $S = MN$ is the total number of draws from all chains. For each scalar summary of interest $\theta$, we compute $B$ and $W$, the between- and within-chain variances:

\begin{align*}
B=\frac{N}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{N} \sum_{n=1}^{N} \theta^{(n m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)} \\
W=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \text { where } s_{m}^{2}=\frac{1}{N-1} \sum_{n=1}^{N}\left(\theta^{(n m)}-\bar{\theta}^{(\cdot m)}\right)^{2}". \text{ Vehtari et al., 2019, p. 5} 
\end{align*}

The between and within chain variances are computed from the chain means of the mids object, that underwent the previous transformations (folding, splitting, rank-normalizing). The average of the chain means per imputation $m$ is $\bar{\theta}^{(\cdot m)}$. The average of these $m$ values is $\bar{\theta}^{(\cdot \cdot)}$. $N$ is the number of iterations ('maxit'). So we compute $B$ as `maxit * var(apply(sims, 2, mean))`. And $W$ is computed as `mean(apply(sims, 2, var))`.  

To get $\widehat{R}$ we need to compute the weighted average of $B$ and $W$, $\widehat{\operatorname{var}}^{+}$, and evaluate this against the within chain variance W. The weighted average is is computed as follows: 

\begin{equation*}
\widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B,
\end{equation*}

Then finally, potential scale reduction factor $\widehat{R}$ is obtained as:

\begin{equation*}
\widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}}.
\end{equation*}

$\widehat{R}$ is computed for each variable with the function 'get.rhat()'. 
```{r warning=FALSE}
# show the contents of the function
get.rhat

```

# Step 3: Apply these functions to get $\widehat{R}$ values

Now we know what's happening, let's apply it.
```{r}
rhat_function(imp, maxit, n.var = n.var)

```

This $\widehat{R}$ value is nice and high, but I also got values below one. Which is theoretically impossible.  

# So my question to you is: How is it possible that we can obtain  $\widehat{R}$ values < 1?